---
title: "DirichletReg -- Article Sources"
author: "Marco J. Maier"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{DirichletReg -- Article Sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
  knitr::opts_chunk$set(
      collapse   = TRUE
    , comment    = "##"
    , tidy       = TRUE
    , fig.width  = 7L
    , fig.height = 7L
    , tidy.opts  = list(blank = TRUE, width.cutoff = 75L, indent = 4L)
)
```

Full R-Code for:

Maier, M. J. (2014). _DirichletReg: Dirichlet Regression for Compositional Data in R._
Research Report Series/Department of Statistics and Mathematics, 125. WU Vienna University of Economics and Business, Vienna.
[doi.org/10.57938/ad3142d3-2fcd-4c37-aec6-8e0bd7d077e1](https://doi.org/10.57938/ad3142d3-2fcd-4c37-aec6-8e0bd7d077e1)

## 4. Application examples

### 4.1. The Arctic lake (common parametrization)

```{r, message = FALSE}
library(DirichletReg)
head(ArcticLake)

AL <- DR_data(ArcticLake[, 1:3])

AL[1:6, ]
```

Code for Fig. 4 (left):
```{r, fig.cap = "Figure 4: Arctic lake: Ternary plot (left) and a scatter plot of depth vs. composition (right)."}
plot(AL, cex = 0.5, a2d = list(colored = FALSE, c.grid = FALSE))
```

Code for Fig. 4 (right):
```{r, fig.cap = "Figure 4: Arctic lake: Ternary plot (left) and a scatter plot of depth vs. composition (right)."}
plot(rep(ArcticLake$depth, 3L), as.numeric(AL), pch = 21L, bg = rep(c("#E495A5", "#86B875", "#7DB0DD"), each = 39L),
  xlab = "Depth (m)", ylab = "Proportion", ylim = c(0.0, 1.0))
```

```{r}
lake1 <- DirichReg(AL ~ depth, ArcticLake)
lake1

coef(lake1)

lake2 <- update(lake1, . ~ . + I(depth^2) | . + I(depth^2) | . + I(depth^2))
anova(lake1, lake2)

summary(lake2)
```




Code for Fig. 5:
```{r, fig.show='hold', fig.cap = "Figure 5: Arctic lake: Fitted values of the quadratic model."}
par(mar = c(4.0, 4.0, 4.0, 4.0) + 0.1)
plot(rep(ArcticLake$depth, 3L), as.numeric(AL), pch = 21L, bg = rep(c("#E495A5", "#86B875", "#7DB0DD"), each = 39L),
  xlab = "Depth (m)", ylab = "Proportion", ylim = c(0.0, 1.0), main = "Sediment Composition in an Arctic Lake")

Xnew <- data.frame(depth = seq(min(ArcticLake$depth), max(ArcticLake$depth), length.out = 100L))
for(i in seq_len(3L)) lines(cbind(Xnew, predict(lake2, Xnew)[, i]), col = c("#E495A5", "#86B875", "#7DB0DD")[i], lwd = 2.0)   
legend("topleft", legend = c("Sand", "Silt", "Clay"), lwd = 2.0, col = c("#E495A5", "#86B875", "#7DB0DD"),
  pt.bg = c("#E495A5", "#86B875", "#7DB0DD"), pch = 21L, bty = "n")

par(new = TRUE)
plot(cbind(Xnew, predict(lake2, Xnew, FALSE, FALSE, TRUE)), lty = "24", type = "l",
  ylim = c(0.0, max(predict(lake2, Xnew, FALSE, FALSE, TRUE))), axes = FALSE, ann = FALSE, lwd = 2.0)
axis(4L)
mtext(expression(paste("Precision (", phi, ")")), 4L, line = 3L)
legend("top", legend = c(expression(hat(mu)[c] == hat(alpha)[c] / hat(alpha)[0]), expression(hat(phi) == hat(alpha)[0])),
  lty = c(1L, 2L), lwd = c(3.0, 2.0), bty = "n")
```

```{r}
AL <- ArcticLake
AL$AL <- DR_data(ArcticLake[, 1:3])

dd <- range(ArcticLake$depth)
X <- data.frame(depth = seq(dd[1L], dd[2L], length.out = 200L))

pp <- predict(DirichReg(AL ~ depth + I(depth^2), AL), X)
```


Code for Fig. 6:
```{r, fig.cap = "Figure 6: Arctic lake: OLS (dashed) vs. Dirichlet regression (solid) predictions."}
plot(AL$AL, cex = 0.1, reset_par = FALSE)
points(toSimplex(AL$AL), pch = 16L, cex = 0.5, col = gray(0.5))

lines(toSimplex(pp), lwd = 3.0, col = c("#6E1D34", "#004E42")[2L])

Dols <- log(cbind(ArcticLake[, 2L] / ArcticLake[, 1L],
                  ArcticLake[, 3L] / ArcticLake[, 1L]))

ols <- lm(Dols ~ depth + I(depth^2), ArcticLake)
p2 <- predict(ols, X)
p2m <- exp(cbind(0.0, p2[, 1L], p2[, 2L])) / rowSums(exp(cbind(0.0, p2[, 1L], p2[, 2L])))

lines(toSimplex(p2m), lwd = 3.0, col = c("#6E1D34", "#004E42")[1L], lty = "21")
```

### 4.2. Blood samples (alternative parametrization)

```{r}
Bld <- BloodSamples
Bld$Smp <- DR_data(Bld[, 1:4])

blood1 <- DirichReg(Smp ~ Disease | 1, Bld, model = "alternative", base = 3L)
blood2 <- DirichReg(Smp ~ Disease | Disease, Bld, model = "alternative", base = 3L)
anova(blood1, blood2)

summary(blood1)
```

Code for Fig. 7:
```{r, fig.cap="Figure 7: Blood samples: Box plots and fitted values (dashed lines indicate the fitted values for each group)."}
par(mfrow = c(1L, 4L), mar = c(4.0, 4.0, 4.0, 2.0) + 0.25)
for(i in 1:4){
  boxplot(Bld$Smp[, i] ~ Bld$Disease, ylim = range(Bld$Smp[, 1:4]), main = paste(names(Bld)[i]),
    xlab = "Disease Type", ylab = "Proportion")
  segments(c(-5.0, 1.5), unique(fitted(blood2)[, i]),
           c( 1.5, 5.0), unique(fitted(blood2)[, i]), lwd = 2.0, lty = 2L)
}
```

```{r}
alpha <- predict(blood2, data.frame(Disease = factor(c("A", "B"))), FALSE, TRUE, FALSE)
L <- sapply(1:2, function(i) ddirichlet(DR_data(Bld[31:36, 1:4]), unlist(alpha[i, ])))
LP <- L / rowSums(L)
dimnames(LP) <- list(paste("C", 1:6), c("A", "B"))
print(data.frame(round(LP * 100, 1), "pred." = as.factor(ifelse(LP[, 1L] > LP[, 2L], "==> A", "==> B"))), print.gap = 2L)
```

Code for Fig. 8:
```{r, fig.cap = "Figure 8: Blood samples: Observed values and predictions."}
B2 <- DR_data(BloodSamples[, c(1L, 2L, 4L)])
plot(B2, cex = 0.001, reset_par = FALSE)

div.col <- colorRampPalette(c("#023FA5", "#c0c0c0", "#8E063B"))(100L)

# expected values
temp <- (alpha / rowSums(alpha))[, c(1L, 2L, 4L)]
points(toSimplex(temp / rowSums(temp)), pch = 22L, bg = div.col[c(1L, 100L)], cex = 2.0, lwd = 0.25)

# known values   
temp <- B2[1:30, ]
points(toSimplex(temp / rowSums(temp)), pch = 21L, bg = (div.col[c(1L, 100L)])[BloodSamples$Disease[1:30]], cex = 0.5, lwd = 0.25)

# unclassified  
temp <- B2[31:36, ]
points(toSimplex(temp / rowSums(temp)), pch = 21L, bg = div.col[round(100 * LP[, 2L], 0L)], cex = 1.0, lwd = 0.5)

legend("topright", bty = "n", legend = c("Disease A", "Disease B", NA, "Expected Values"), pch = c(21L, 21L, NA, 22L),
  pt.bg = c(div.col[c(1L, 100L)], NA, "white"))
```

### 4.3. Reading skills data (alternative parametrization)

```{r}
RS <- ReadingSkills
RS$acc <- DR_data(RS$accuracy)

RS$dyslexia <- C(RS$dyslexia, treatment)
rs1 <- DirichReg(acc ~ dyslexia * iq | dyslexia * iq, RS, model = "alternative")
rs2 <- DirichReg(acc ~ dyslexia * iq | dyslexia + iq, RS, model = "alternative")
anova(rs1, rs2)
```

Code for Fig. 9:
```{r, fig.cap="Figure 9: Reading skills: Predicted values of Dirichlet regression and OLS regression."}
g.ind <- as.numeric(RS$dyslexia)
g1 <- g.ind == 1L   # normal
g2 <- g.ind != 1L   # dyslexia
par(mar = c(4.0, 4.0, 4.0, 4.0) + 0.25)
plot(accuracy ~ iq, RS, pch = 21L, bg = c("#E495A5", "#39BEB1")[3L - g.ind],
cex = 1.5, main = "Dyslexic (Red) vs. Control (Green) Group",
xlab = "IQ Score", ylab = "Reading Accuracy", xlim = range(ReadingSkills$iq))

x1 <- seq(min(RS$iq[g1]), max(RS$iq[g1]), length.out = 200L)
x2 <- seq(min(RS$iq[g2]), max(RS$iq[g2]), length.out = 200L)
n <- length(x1)
X <- data.frame(dyslexia = factor(rep(0:1, each = n), levels = 0:1, labels = c("no", "yes")), iq = c(x1, x2))
pv <- predict(rs2, X, TRUE, TRUE, TRUE)

lines(x1, pv$mu[1:n, 2L], col = c("#E495A5", "#39BEB1")[2L], lwd = 3.0)
lines(x2, pv$mu[(n + 1L):(2L * n), 2L], col = c("#E495A5", "#39BEB1")[1L], lwd = 3.0)

a <- RS$accuracy
logRa_a <- log(a / (1.0 - a))
rlr <- lm(logRa_a ~ dyslexia * iq, RS)
ols <- 1 / (1 + exp(-predict(rlr, X)))
lines(x1, ols[1:n], col = c("#AD6071", "#00897D")[2L], lwd = 3.0, lty = 2L)
lines(x2, ols[(n + 1L):(2L * n)], col = c("#AD6071", "#00897D")[1L], lwd = 3.0, lty = 2L)

### precision plot
par(new = TRUE)
plot(x1, pv$phi[1:n], col = c("#6E1D34", "#004E42")[2L], lty = "11", type = "l", ylim = c(0.0, max(pv$phi)), axes = FALSE,
  ann = FALSE, lwd = 2.0, xlim = range(RS$iq))
lines(x2, pv$phi[(n + 1L):(2L * n)], col = c("#6E1D34", "#004E42")[1L], lty = "11", type = "l", lwd = 2.0)
axis(4L)
mtext(expression(paste("Precision (", phi, ")")), 4L, line = 3)
legend("topleft", legend = c(expression(hat(mu)), expression(hat(phi)), "OLS"), lty = c(1L, 3L, 2L), lwd = c(3.0, 2.0, 3.0),
  bty = "n")
```

```{r}
a <- RS$accuracy
logRa_a <- log(a / (1 - a))
rlr <- lm(logRa_a ~ dyslexia * iq, RS)
summary(rlr)

summary(rs2)

confint(rs2)

confint(rs2, exp = TRUE)
```

Code for 10:
```{r, fig.height=7/2*3, fig.cap="Figure 10: Reading skills: Various residual plots of OLS and Dirichlet regression models. The left columns shows OLS, the right column Dirichlet models. In the first row, QQ-plots show deviation from a normal distribution of residuals. The second row plots residuals against the covariate IQ. The third row plots residuals against fitted values."}
gcol <- c("#E495A5", "#39BEB1")[3L - as.numeric(RS$dyslexia)]
tmt <- c(-3.0, 3.0)

par(mfrow = c(3L, 2L), cex = 0.8)

qqnorm(residuals(rlr, "pearson"), ylim = tmt, xlim = tmt, pch = 21L, bg = gcol,
  main = "Normal Q-Q-Plot: OLS Residuals", cex = 0.75, lwd = 0.5)
abline(0.0, 1.0, lwd = 2L)
qqline(residuals(rlr, "pearson"), lty = 2L)

qqnorm(residuals(rs2, "standardized")[, 2L], ylim = tmt, xlim = tmt, pch = 21L, bg = gcol,
  main = "Normal Q-Q-Plot: DirichReg Residuals", cex = 0.75, lwd = 0.5)
abline(0.0, 1.0, lwd = 2L)
qqline(residuals(rs2, "standardized")[, 2L], lty = 2L)

plot(ReadingSkills$iq, residuals(rlr, "pearson"), pch = 21L, bg = gcol, ylim = tmt,
  main = "OLS Residuals", xlab = "IQ", ylab = "Pearson Residuals", cex = 0.75, lwd = 0.5)
abline(h = 0.0, lty = 2L)
plot(ReadingSkills$iq, residuals(rs2, "standardized")[, 2L], pch = 21L, bg = gcol, ylim = tmt,
  main = "DirichReg Residuals", xlab = "IQ", ylab = "Standardized Residuals", cex = 0.75, lwd = 0.5)
abline(h = 0.0, lty = 2L)

plot(fitted(rlr), residuals(rlr, "pearson"), pch = 21L, bg = gcol, ylim = tmt,
  main = "OLS Residuals", xlab = "Fitted", ylab = "Pearson Residuals", cex = 0.75, lwd = 0.5)
abline(h = 0.0, lty = 2L)
plot(fitted(rs2)[, 2L], residuals(rs2, "standardized")[, 2L], pch = 21L, bg = gcol, ylim = tmt,
  main = "DirichReg Residuals", xlab = "Fitted", ylab = "Standardized Residuals", cex = 0.75, lwd = 0.5)
abline(h = 0.0, lty = 2L)
```
